Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-11-22 concurrency/CPP 2024/paper.tex
initial revise

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Duc Than Nguyen}
%\authornote{Both authors contributed equally to this research.}
\email{dnguye96@uic.edu}
\orcid{}
\affiliation{%
  \institution{University of Illinois Chicago}
  \streetaddress{}
  \city{Chicago}
  \state{Illinois}
  \country{USA}
  \postcode{}
}

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ \\
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
\end{mathpar}

\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

@@

\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}
	\end{mathpar}
	\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
\end{mathpar}

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
		\right\rangle
	}\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto \left(n', n'\right) \ \ast \\ \nodeboxrep\left(n'\right) \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
			\\ 
			(\mathsf{if} \ \mathit{res} \ \\ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right))
		\end{array}
		\ \Bigg | \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ m$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The desired specification of \texttt{insert} is:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\  \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{\color{blue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
		\end{array}
		\right\rangle
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
			\\ 
			(\mathsf{if} \ \mathit{res} \\ 
			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ 
			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
		\end{array}
		\Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}. Our definition of $\treerep$ for the give-up template closely follows that of Krishna et al.

@@

Compared to prior template proofs in Iris and GRASShopper, our proof development is significantly larger (although direct comparison is difficult, since GRASShopper is an automatic prover). The biggest structural difference is that we need to directly state loop invariants for our imperative code, while the Iris proofs handle recursion with a simpler Löb-induction approach. Coming up with the loop invariants was the hardest part of the proofs. Our proofs also deal with the details of real C code: proving absence of integer overflows, accurately representing struct-and-pointer-based data structures instead of functional-programming-style structured data, etc. In exchange for this extra effort, the code we verify constitutes complete C programs that can be compiled and executed.
The full development of our mechanization effort is available online at [link redacted for anonymity].
%\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.

@@

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
	\end{array}
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure*}

\begin{figure*}[!ht]
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the lock-coupling \texttt{insert} function}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the give-up \texttt{insert} function}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{Proof outlines for the \texttt{insert} function}
	\label{proof_lock_giveup_insert}
\end{figure*}

>>>>

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Duc Than Nguyen}
%\authornote{Both authors contributed equally to this research.}
\email{dnguye96@uic.edu}
\orcid{https://orcid.org/0000-0002-6810-897X}
\affiliation{%
  \institution{University of Illinois Chicago}
  \streetaddress{}
  \city{Chicago}
  \state{IL}
  \country{USA}
  \postcode{}
}
\author{Lennart Beringer}
%\authornote{Both authors contributed equally to this research.}
\email{eberinge@cs.princeton.edu}
\orcid{https://orcid.org/0000-0002-1570-3492}
\affiliation{%
	\institution{Princeton University}
	\streetaddress{}
	\city{Princeton}
	\state{NJ}
	\country{USA}
	\postcode{}
}

\author{William Mansky}
%\authornote{Both authors contributed equally to this research.}
\email{mansky1@uic.edu}
\orcid{https://orcid.org/0000-0002-5351-895X}
\affiliation{%
	\institution{University of Illinois Chicago}
	\streetaddress{}
	\city{Chicago}
	\state{IL}
	\country{USA}
	\postcode{}
}

\author{Shengyi Wang}
%\authornote{Both authors contributed equally to this research.}
\email{shengyiw@princeton.edu}
\orcid{https://orcid.org/0000-0002-2286-8703}
\affiliation{%
	\institution{Princeton University}
	\streetaddress{}
	\city{Princeton}
	\state{NJ}
	\country{USA}
	\postcode{}
}

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ \\
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m) \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
\end{mathpar}

\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

@@

\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
	\end{mathpar}
	\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
\end{mathpar}

@@

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
		\right\rangle
	}\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto \left(n', n'\right) \ \ast \\ \nodeboxrep\left(n'\right) \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
			\\ 
			(\mathsf{if} \ \mathit{res} \ \\ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right))
		\end{array}
		\ \Bigg | \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ (m)$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The desired specification of \texttt{insert} is:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\  \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{\color{blue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ (m)
		\end{array}
		\right\rangle
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
			\\ 
			(\mathsf{if} \ \mathit{res} \\ 
			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ 
			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
		\end{array}
		\Bigg| \ \treerep\ (m) \
		\right\rangle
	}
\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}. Our definition of $\treerep$ for the give-up template closely follows that of Krishna et al.

@@

Compared to prior template proofs in Iris and GRASShopper, our proof development is significantly larger (although direct comparison is difficult, since GRASShopper is an automatic prover). The biggest structural difference is that we need to directly state loop invariants for our imperative code, while the Iris proofs handle recursion with a simpler Löb-induction approach. Coming up with the loop invariants was the hardest part of the proofs. Our proofs also deal with the details of real C code: proving absence of integer overflows, accurately representing struct-and-pointer-based data structures instead of functional-programming-style structured data, etc. In exchange for this extra effort, the code we verify constitutes complete C programs that can be compiled and executed.
The full development of our mechanization effort is available online at \href{https://zenodo.org/record/8337004}{\color{blue}{https://zenodo.org/record/8337004}}.

%[link redacted for anonymity].

%\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.

@@

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ (m)
	\end{array}
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure*}

\begin{figure*}[!ht]
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the lock-coupling \texttt{insert} function}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{Proof outline of the give-up \texttt{insert} function}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{Proof outlines for the \texttt{insert} function}
	\label{proof_lock_giveup_insert}
\end{figure*}
